---
title: "NORS National Outbreak Data Analysis and Visualization"
author: "Carmen Raterman"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
---

# To retrieve the data:

1\. Visit the CDC NORS website at: <https://wwwn.cdc.gov/norsdashboard/>

2\. Check only the "Foodborne" box for type of outbreak

3\. Under "Year" select "Custom range" and input from 2011 to 2021 (10 years)

4\. Scroll down to "Download Data" and click "Download current search data (excel)" to import the data on your local machine

5\. Follow the steps below for importing the data to R Studio and performing analysis and visualization

# Import Data

```{r data-setup}
options(repos = c(CRAN = "https://cloud.r-project.org"))

#necessary libraries to download and call
install.packages("pacman")
library(pacman)
p_load(tidyverse, rio, ggplot2, readxl)

#read in the data from excel format to R compatible format
nat_outbreak_data <- read_excel("NationalOutbreakPublicDataTool.xlsx")

#this function allows us to inspect the first few observations of the data to double check that it imported successfully
head(nat_outbreak_data)

```

# Data Visualization and Analysis

## Poisson GLM

```{r, analysis}

#re-arranging the data to aggregate for analysis, and dropping "multi-state"
#grouping the observations by state and year to get a count of total illnesses by state and year
outbreak_data_agg <- nat_outbreak_data %>%
  filter(State != "Multistate") %>%
  group_by(State, Year) %>%
  summarize(Total_Illnesses = sum(Illnesses))


#now doing poisson glm 
model_output <- glm(Total_Illnesses ~ State + Year, family = "poisson", data = outbreak_data_agg)
summary(model_output)

```

```{r, map-creation}
library(sf)
library(tidyverse)
library(tmap)
library(tmaptools)

#read in the shapefile of states
states <- st_read("States_shapefile.shp")

#re-arrange our aggregate data to now show prevalence
#doing so normalizes the data a bit more for effective visualization on the map
outbreak_data_agg_prev <- outbreak_data_agg %>% 
  group_by(Year) %>% 
  mutate(sumYear = sum(Total_Illnesses)) %>% 
  ungroup() %>% 
  mutate(prevYear = (Total_Illnesses/sumYear) * 100)

#pivot wider
illness_wide <- outbreak_data_agg_prev %>%
                select(State, prevYear, Year) %>% 
                pivot_wider(values_from = "prevYear", names_from = "Year") %>% 
                mutate(State = toupper(State))

#left joining so we can combine the illness data to the shapefile for mapping
state <- states %>%  
          left_join(illness_wide, by=c("State_Name" = "State"))

#create map for 2011 and store it as variable for side-by-side comparison below
map_2011 <- tm_shape(state) +
  tm_fill(col="2011") +
  tm_borders() +
  tm_layout(title = "2011 Prevalence of All Foodborne \nIllnesses Reported to NORS in the US",
            legend.position = c(0.1, 0.1),
            legend.bg.alpha = 0.5,
            legend.text.size = 1.0,
            title.size = 1.1,
            title.position = c(0.45, 0.9))

#display the individual map for 2011
tm_shape(state) +
  tm_fill(col="2011") +
  tm_borders() +
  tm_layout(title = "2011 Prevalence of All Foodborne \nIllnesses Reported to NORS in the US",
            legend.position = c(0.1, 0.1),
            legend.bg.alpha = 0.5,
            legend.text.size = 1.0,
            title.size = 1.1,
            title.position = c(0.45, 0.9))

#create map for 2021 and store it as variable for side-by-side comparison below
map_2021 <- tm_shape(state) +
  tm_fill(col="2021") +
  tm_borders() +
  tm_layout(title = "2021 Prevalence of All Foodborne \nIllnesses Reported to NORS in the US",
            legend.position = c(0.1, 0.1),
            legend.bg.alpha = 0.5,
            legend.text.size = 1.0,
            title.size = 1.1,
            title.position = c(0.45, 0.9))

#display the individual map for 2021
tm_shape(state) +
  tm_fill(col="2021") +
  tm_borders() +
  tm_layout(title = "2021 Prevalence of All Foodborne \nIllnesses Reported to NORS in the US",
            legend.position = c(0.1, 0.1),
            legend.bg.alpha = 0.5,
            legend.text.size = 1.0,
            title.size = 1.1,
            title.position = c(0.45, 0.9))

#arrange the two maps together for visualization across 10 years
tmap_arrange(map_2011, map_2021)

```
